Type theories as quotient inductive-recursive types

### 2025-01-22 tex/main.tex
Draft S3.2

\usepackage[obeyFinal]{todonotes}
\usepackage{manfnt}
\newcommand{\danger}{\marginpar[\hfill\dbend]{\dbend\hfill}}
\newcommand{\LT}[2][]{\todo[inline,author={L-T},caption={},#1]{#2}}

@@

\begin{abstract}
  Type theory can be defined in a type theory as a quotient inductive-inductive type, but well-typed terms are littered with explicit coercions, i.e.\ \transp's, along equality constructors and proofs need to take account of coercions painstakingly alongside the interesting part.
  This mess has been dubbed the `transport hell' but typically suppressed in presentation for clarity, hiding the gap between the intention and the formalisation of type theory in type theory.
  In this paper, we aim for shortening the gap using quotient inductive-inductive-recursive types and definitions by overlapping patterns altogether, reducing the use of \transp's.
  As a case study, we investigate (parallel) substitution calculus and type theory with Coquand universes and $\Pi$-types as quotient inductive-inductive-recursive types and compare ours with quotient inductive-inductive definitions.
\end{abstract}

@@

\paragraph*{Contributions}
\begin{itemize}
  \item Exploration of the use of quotient inductive-inductive-recursive types and definitions by overlapping patterns.
    In particular, we give in a type theory:
    \begin{itemize}
      \item a definition of parallel substitution calculus;
      \item a definition of type theory with Coquand universes and $\Pi$-types,
    \end{itemize}
   using quotient inductive-inductive-recursive types and definitions by overlapping patterns to define type substitution and term substitution partially.
  \item Comparison with other definitions using quotient inductive-inductive types.
\end{itemize}

@@

\section{Metatheory and formalisation}
\cite{UFP2013}
\LT{
Points to include:
\begin{enumerate}
  \item identity type $x =^{A} y$ for $x, y : A$, dependent identity type, $t =^{P}_{p} u \defeq \transfib{P}{p}{t} =^{P y} u$ for $t : P(x)$ and $u : P(y)$, heterogeneous equality~\cite{McBride1999} in \Agda $x \simeq y$ for $x : A$ and $y : B$
  \item We work with intensional type theory with uniqueness of identity proof and function extensionality (only used for NbE and the standard model for the extensional identity type).
  \item We use Agda to formally implement our definitions with the following options: \texttt{-{}-with-K}, \texttt{-{}-local-confluence-check}, \texttt{-{}-exact-split}, and \texttt{-{}-rewriting} using postulated equations~\cite{Licata2011} to introduce equality constructors for quotient types.
\end{enumerate}}

\subsection{Inductive-recursive types}
\cite{Dybjer2003,Dybjer2000,Dybjer1999}

@@

\subsection{Definitions by overlapping patterns} \label{sec:meta:overlapping}
\cite{Cockx2014,Altenkirch2016a}
\subsection{Quotient inductive(-inductive)-recursive types}
\LT{Give an example of QIRT: join list?}
\subsection{Definitions by rewrite rules}
\cite{Cockx2020,Cockx2021}

@@

Type substitution takes a substitution $\sigma : \Sub\;\Gamma\;\Delta$ and a type under $\Delta$ to form a type under $\Gamma$, so type substitution as a constructor has the type
\begin{alignat*}{3}
  [\blank]\blank &: \implicit{\Gamma, \Delta} \; \Sub\;\Gamma\;\Delta \to \Ty\;\Delta \to \Ty\;\Gamma.
\end{alignat*}
where $\Gamma$ and $\Delta$ (coloured {\color{gray}grey}) above are quantified implicitly.

If we were defining cwfs, type substitution would be enough for $\Ty$.
However, we need a base case in $\Ty$ for any context $\Gamma$ for our inductive definition (i.e.\ the initial cwf), 
otherwise the type $\Ty$ would be empty as well as other types.
Hence another constructor is introduced:
\begin{alignat*}{3}
  \UU & : \implicit{\Gamma} & \Ty\, \Gamma.
\end{alignat*}
For now, $\UU$ serves as a constant in $\Ty$, but we will reuse $\UU$ for the type of small types later.

Substitutions from $\Gamma$ to $\Delta$ can be understood intuitively as lists of terms of type $A$ under the context $\Gamma$ for each $A$ in $\Delta$, so we have the empty substitution $\emptysub$ and substitutions $\sigma, t$ extended by some term~$t$:
\begin{alignat*}{3}
  \emptysub & : \implicit{\Gamma}\;\Sub\;\Gamma\;\emptyctx \\
  \blank,\blank & : \implicit{\Gamma, \Delta, A}\;(\sigma : \Sub\,\Gamma\,\Delta) \to \Tm\;\Gamma\;([ \sigma ]\, A) \to \Sub\;\Gamma\;(\Delta, A),
\end{alignat*}
Note that type substitution $[\sigma]\;A$ is needed, because $A$ is well-formed under the context $\Delta$ instead of $\Gamma$.
As substitution calculus is closely related to cwfs (more precisely, it is the initial model), substitutions also have the identity substitution $\idS$ and composition $\blank;\blank$
\begin{alignat*}{3}
  \idS & : \implicit{\Gamma}\;\Sub\;\Gamma\;\Gamma \\
  \blank;\blank & : \implicit{\Gamma, \Delta, \Theta}\;\Sub\;\Gamma\;\Delta \to \Sub\;\Delta\;\Theta \to \Sub\;\Gamma\;\Theta \\
\end{alignat*}
satisfying certain laws introduced later.
Context comprehension is given by projections from \emph{non-empty} substitutions $\Sub\;\Gamma\;(\Delta, A)$:
\begin{alignat*}{3}
  \pi_1 & : \implicit{\Gamma, \Delta, A}\;\Sub\;\Gamma\;(\Delta, A) \to \Sub\;\Gamma\;\Delta \\
  \pi_2 & : \implicit{\Gamma, \Delta, A}\;(\sigma : \Sub\;\Gamma\;(\Delta, A)) \to \Tm\;\Gamma\;([ \pi_1\,\sigma ]\; A)
\end{alignat*}
where $\pi_1\,\sigma$ and $\pi_2\,\sigma$ can be intuitively understood as the \emph{tail} and the \emph{head} of a non-empty substitution $\sigma : \Sub\;\Gamma\;(\Delta, A)$ respectively.
Again, a type substitution $[\pi_1\,\sigma]\,A$ is needed for $\pi_2$, because $A$ is under the context $\Delta$ instead of $\Gamma$.
Finally, we also have term substitution $[\sigma]\,t$ whose type is $[\sigma]\,A$ for a term $t : \Tm\;\Delta\;A$
\begin{alignat*}{3}
  [\blank] \blank & : \implicit{\Gamma,\Delta, A}\;(\sigma : \Sub\;\Gamma\;\Delta) \to \Tm\;\Delta\;A \to \Tm\;\Gamma\;([\sigma]\; A)
\end{alignat*}
where the symbol $[\blank]\blank$ is overloaded.

@@

Laws for substitutions are stipulated using equality constructors.
\begin{alignat*}{5}
  \mathsf{idr}    & : \implicit{\Gamma, \Delta, \sigma} && {\sigma ; \idS_{\Delta}} && =^{\Sub\,\Gamma\,\Delta} && {\sigma} \\
  \mathsf{idl}    & : \implicit{\Gamma, \Delta, \sigma} && {\idS_{\Gamma} ; \sigma} && =^{\Sub\,\Gamma\,\Delta} && {\sigma} \\
  ;\text{-}\mathsf{assoc} & : \implicit{\Gamma, \Delta, \Theta, \Xi, \sigma, \tau, \gamma} && (\sigma ; \tau) ; \gamma && =^{\Sub\,\Gamma\,\Theta} &&  \sigma ; (\tau ; \gamma) \\
  \mathsf{concat} & : \implicit{\Gamma, \Delta, \Theta, \sigma, \tau, A, t} && \sigma ; (\tau , t)      && =^{\Sub\,\Gamma\,(\Theta, A)} &&  (\sigma ; \tau) , \alert{\transfib{\Tm\,\Gamma}{[;]_{\Ty}^{-1}}{\color{black}[ \sigma ] t}} \\
  \emptyctx\eta   & : \implicit{\Gamma, A, \sigma} && \sigma                   && =^{\Sub\,\Gamma\,\emptyctx} & \emptysub \\
  \pi\eta         & : \implicit{\Gamma, \Delta, \sigma} && \sigma                   && =^{\Sub\,\Gamma\,(\Delta, A)} &&  (\pi_1 \sigma, \pi_2 \sigma) \\
\end{alignat*}
In particular, $\blank;\blank$ acts like a list concatenation, so conceptually $\sigma; (\tau, t)$ is equal to $(\sigma; \tau), t$ whereas the term $t$ is of type $\Tm\;\Delta\;([\tau]\;A)$. 
Moving the term $t$ from the left-hand side to the right-hand side requires us to apply substitution $\sigma$ to the term $t$ to obtain an inhabitant of $\Tm\;\Gamma\;[\sigma][\tau]\,A$.
By definition of the substitution extension, however, $[\sigma]\;t$ requires to be $\Tm\;\Gamma\;[\sigma; \tau]\,A$ instead.
Therefore, we need to transport $[\sigma]\;t$ explicitly along $[;]_{\Ty}^{-1}\colon [\sigma]\,[\tau]\;A = [\sigma;\tau]\,A$ to obtain a term of type $\Tm\;\Gamma\;([\sigma;\tau]\;A)$.

@@

Similarly, for term substitution, terms are transported along the corresponding rules:
\begin{alignat*}{5}
  [\idS]_t         & : \implicit{\Gamma, A, t} && {[\,\idS\,]\,t}         && =^{\Tm\;\Gamma}_{\alert{[\idS]_\Ty}}  && t \\
  [;]_t            & : \implicit{\Gamma, \Delta, \Theta, \sigma, \tau, t} && {[\,\sigma ; \tau\,]\,t} && =^{\Tm\;\Gamma}_{\alert{[;]_{\Ty}}}   && {[ \sigma ]\,[ \tau ]\,t}
\end{alignat*}
equivalent to the following (homogeneous) identities
\[
  \transfib{\Tm\;\Gamma}{[\idS]_{\Ty}}{[\idS]\,t} =^{\Tm\,\Gamma\,A} t
  \quad\text{and}\quad
  \transfib{\Tm\;\Gamma}{[;]_{\Ty}}{[\sigma;\tau]\,t} =^{\Tm\,\Gamma\,([\sigma]\,[\tau]A)} [\sigma]\,[\tau]\,t
\]
respectively.
Constructors introduced so far complete the definition of substitution calculus.

@@

To retain the intuitive way of reasoning such as \eqref{eq:pi2-comp-proof} \emph{formally}, we make type substitution \emph{definitional}, since the root cause is the type mismatch after type substitution.

One possibility is to define substitution calculus as a quotient inductive-inductive-\emph{recursive} type whereas type substitution as a simultaneous recursive function should compute.
Now that $[ \idS ]_{\Ty}\;A = A$ and $[ \sigma ; \tau ]_{\Ty}\;A = [ \sigma ]_{\Ty}\;([ \tau ]_{\Ty}\;A)$ along do not make $[\blank]_{\Ty}\blank$ a function, we may define it by
\begin{alignat*}{3}
[\blank]_{\Ty} \blank &: \implicit{\Gamma, \Delta} \; \Sub\;\Gamma\;\Delta \to \Ty\;\Delta \to \Ty\;\Gamma \\
[ \sigma ]_{\Ty}\;\UU & = \UU
\end{alignat*}
However, this definition is \emph{not} inductive-recursive in the sense of Dybjer and Setzer~\cite{Dybjer2000,Dybjer2003}: in their schema, the codomain of a simultaneous recursive function does not refer to any inductive type being defined.
While its general schema remains underdeveloped, this definition is accepted in proof assistants such as \Agda and used in other formalisation of type theory~\cite{Danielsson2006}.
Still, although $[\idS]_{\Ty}\,A$ and $[\sigma; \tau]_{\Ty}\,A$ can be proved equal to $A$ and $[\sigma]_{\Ty}\,[\tau]_{\Ty}\,A$ respectively, they are not definitional but propositional.
This definition fails to eliminate any transport in the definition of substitution calculus.

Our second attempt is to define type substitution with additional rules:
\begin{alignat}{3}
[\blank]_{\Ty} \blank &: \implicit{\Gamma, \Delta} \; \Sub\;\Gamma\;\Delta \to \Ty\;\Delta \to \Ty\;\Gamma \nonumber \\
[ \sigma ]_{\Ty}       \;\UU  & = \UU                                   \label{eq:type-sub-at2-1} \\
[ \idS ]_{\Ty}         \;A    & = A                                     \label{eq:type-sub-at2-2} \\
[ \sigma ; \tau ]_{\Ty}\;A    & = [ \sigma ]_{\Ty}\;([ \tau ]_{\Ty}\;A) \label{eq:type-sub-at2-3}
\end{alignat}
However, this definition does \emph{not} compute \eqref{eq:type-sub-at2-2} or \eqref{eq:type-sub-at2-3} definitionally in a type-theoretic proof assistant.
By design, definitions by pattern matching will be translated to a form that corresponds to the \emph{eliminator} either using first-match semantics~\cite{Cockx2020a} or an explicit construct for pattern matching, so only the eliminator for $\Ty\;\Delta$ will be considered in the above definition.
\LT[noinline]{How about \Coq and \Lean?}
As a result, the above definition amounts to the following definitional clause:
\begin{alignat*}{3}
[\blank]_{\Ty} \blank &: \implicit{\Gamma, \Delta} \; \Sub\;\Gamma\;\Delta \to \Ty\;\Delta \to \Ty\;\Gamma \\
[ \sigma ]_{\Ty}\;\UU  & \judgeq \UU
\end{alignat*}
where $\judgeq$ is used to emphasise the definitional equality: we have done nothing different from our first attempt! 
%As a result, neither  are definitional.
We may swap the order of function clauses to make \eqref{eq:type-sub-at2-2} and \eqref{eq:type-sub-at2-3} definitional, but then \eqref{eq:type-sub-at2-1} will be translated to 
\begin{alignat*}{3}
[ \emptysub ]_{\Ty}\;\UU         & \judgeq \UU \\
[ \sigma, t ]_{\Ty}\;\UU         & \judgeq \UU \\
[ \pi_1\,\sigma ]_{\Ty}\;\UU     & \judgeq \UU,
\end{alignat*}
making $[\sigma]\;\UU = \UU$ propositional instead.
Not being able to compute $[ \sigma ]_{\Ty}\;\UU = \UU$ definitionally would make our later extension with a type $\El\,A$ of elements of $A$ require another transport in the definition.
Likewise, for any extension with a type former such as $\Pi$-types, structural rules would still be propositional.
That is, using the standard semantics of function definition, we are not able to make type substitution definitional on \emph{both} $\Sub\;\Gamma\;\Delta$ and $\Ty\;\Delta$.

To ensure that type substitution is definitional for all \emph{structural rules}, we turn to a definition by overlapping patterns (\Cref{sec:meta:overlapping}) instead to make each clause definitional. 
In addition, we can include other derived rules such as $[\pi_1(\sigma, t)]_{\Ty}\;A = [ \sigma]_{\Ty}\;A$ as long as it is locally confluent and terminating.
We then end up with the following definition:
\begin{alignat}{3}
[\blank]_{\Ty} \blank            &: \implicit{\Gamma, \Delta} \; \Sub\,\Gamma\,\Delta \to \Ty\,\Delta \to \Ty\,\Gamma \nonumber \\
[ \idS ]_{\Ty}\;A                & \jdeq A \label{def:definitional-type-subst-1}\\
[ \sigma ; \tau ]_{\Ty}\;A       & \jdeq [ \sigma ]_{\Ty}\;([ \tau ]_{\Ty}\;A) \label{def:definitional-type-subst-2}\\
[ \pi_1(\sigma, t) ]_{\Ty}\;A    & \jdeq [\sigma]_\Ty\;A                    \label{def:definitional-type-subst-3} \\
[ \pi_1(\sigma; \tau) ]_{\Ty}\;A & \jdeq [\sigma]_\Ty\;([\pi_1\tau]_\Ty\;A) \label{def:definitional-type-subst-4} \\
[ \sigma ]_{\Ty}\;\UU            & \jdeq \UU                                \label{def:definitional-type-subst-5}
\end{alignat}
We first check that this definition of type substitution is valid.\footnote{%
\Agda implements a local confluence checker with the experimental option \text{-{}-local-confluence-check} which is turned on globally in our formalisation, but we do not trust experimental features blindly.}
\begin{proposition}[Local confluence]
  Type substitution $[\blank]_{\Ty}\blank$ is locally confluent.
  \danger
\end{proposition}
\begin{proof}
We consider each combination of \eqref{def:definitional-type-subst-5} with \eqref{def:definitional-type-subst-1}--\eqref{def:definitional-type-subst-4}, that is, 
  \[
    [\idS]_{\Ty}\;\UU, \quad [\sigma;\tau]_{\Ty}\;\UU, \quad [\pi_1(\sigma, t)]_{\Ty}\;\UU, \quad\text{and}\quad [\pi_1(\sigma; \tau)]_{\Ty}\;\U
  \]
which are all local peaks, and show that each applicable clause does result in the same term.
For example, $[\sigma;\tau]_{\Ty}\;\UU \judgeq [\sigma]_{\Ty}\;([\tau]_{\Ty}\;\UU) \judgeq [\sigma]_{\Ty}\;\UU \judgeq \UU$ using \eqref{def:definitional-type-subst-2} as the first reduction and $[\sigma;\tau]_{\Ty}\;\UU \judgeq \UU$ using \eqref{def:definitional-type-subst-5} as the first reduction.
It is easy to see that every local peak reduces to the same term $\UU$, so $[\blank]_{\Ty}\blank$ is locally confluent.

>>>>

\usepackage[obeyFinal,textsize=footnotesize]{todonotes}
\usepackage{manfnt}
\newcommand{\danger}{\marginpar[\hfill\dbend]{\dbend\hfill}}
\newcommand{\LT}[2][]{\todo[inline,author={L-T},caption={},#1]{#2}}

@@

\begin{abstract}
  Type theory can be defined in a type theory as a quotient inductive-inductive type, but well-typed terms are littered with explicit coercions, i.e.\ \transp's, along equality constructors and proofs need to take account of coercions painstakingly alongside the interesting part.
  This mess has been dubbed the `transport hell' but typically suppressed in presentation for clarity, hiding the gap between the intention and the formalisation of type theory in type theory.
  In this paper, we aim for shortening the gap using quotient inductive-inductive-recursive types and definitions by overlapping patterns altogether, reducing the use of \transp's.
  As a case study, we investigate (parallel) substitution calculus and type theory with a universe and $\Pi$-types as quotient inductive-inductive-recursive types and compare ours with quotient inductive-inductive definitions.
\end{abstract}

@@

\paragraph*{Contributions}
\begin{itemize}
  \item Exploration of the use of quotient inductive-inductive-recursive types and definitions by overlapping patterns.
    In particular, we give in a type theory:
    \begin{itemize}
      \item a definition of parallel substitution calculus;
      \item a definition of type theory with a universe and $\Pi$-types,
    \end{itemize}
   using quotient inductive-inductive-recursive types and definitions by overlapping patterns to define type substitution and term substitution partially.
  \item Comparison with other definitions using quotient inductive-inductive types.
\end{itemize}

@@

\section{Metatheory and formalisation}

\cite{UFP2013}
\LT{
Points to include:
\begin{enumerate}
  \item identity type $x =^{A} y$ for $x, y : A$, dependent identity type, $t =^{P}_{p} u \defeq \transfib{P}{p}{t} =^{P y} u$ for $t : P(x)$ and $u : P(y)$, heterogeneous equality~\cite{McBride1999} in \Agda $x \simeq y$ for $x : A$ and $y : B$
  \item We work with intensional type theory with uniqueness of identity proof and function extensionality (only used for NbE and the standard model for the extensional identity type).
  \item We use Agda to formally implement our definitions with the following options: \texttt{-{}-with-K}, \texttt{-{}-local-confluence-check}, \texttt{-{}-exact-split}, and \texttt{-{}-rewriting} using postulated equations~\cite{Licata2011} to introduce equality constructors for quotient types.
\end{enumerate}}

\LT{The symbol \textdbend indicates a property that is proved informally or checked partially with \Agda (because of postulated quotient inductive types).}
\subsection{Inductive-recursive types}
\cite{Dybjer2003,Dybjer2000,Dybjer1999}

@@

\subsection{Definitions by overlapping patterns} \label{sec:meta:overlapping}
\cite{Cockx2014,Altenkirch2016a}
\LT{We use $f\;\vec{x} \reduce \vec{t}$ to indicate a function clause of a definition by overlapping patters.}
\subsection{Quotient inductive(-inductive)-recursive types}
\LT{Give an example of QIRT: join list?}
\subsection{Definitions by rewrite rules}
\cite{Cockx2020,Cockx2021}

@@

Type substitution takes a substitution $\sigma : \Sub\;\Gamma\;\Delta$ and an inhabitant $A : \Ty\;\Delta$ to form a type under $\Gamma$, so type substitution as a constructor has the type
\begin{alignat*}{3}
  [\blank]\blank &: \implicit{\Gamma, \Delta} \; \Sub\;\Gamma\;\Delta \to \Ty\;\Delta \to \Ty\;\Gamma.
\end{alignat*}
where $\Gamma$ and $\Delta$ (coloured {\color{gray}grey}) above are quantified implicitly.

If we were defining the notion of cwfs, type substitution would be enough for $\Ty$.
For our inductive definition (i.e.\ the initial cwf), however, we need a base case in $\Ty\;\Gamma$, otherwise the type $\Ty$ would be empty as well as other types.
Hence another constructor is introduced:
\begin{alignat*}{3}
  \UU & : \implicit{\Gamma} & \Ty\, \Gamma.
\end{alignat*}
For now, $\UU$ serves as a constant, but we will reuse $\UU$ for the type of small types later.

Substitutions from $\Gamma$ to $\Delta$ can be understood intuitively as lists of terms of type $A$ under the context $\Gamma$ for each $A$ in $\Delta$, so we have the empty substitution $\emptysub$ and substitution extensions $\sigma, t$ by some term~$t$:
\begin{alignat*}{3}
  \emptysub & : \implicit{\Gamma}\;\Sub\;\Gamma\;\emptyctx \\
  \blank,\blank & : \implicit{\Gamma, \Delta, A}\;(\sigma : \Sub\,\Gamma\,\Delta) \to \Tm\;\Gamma\;([ \sigma ]\, A) \to \Sub\;\Gamma\;(\Delta, A),
\end{alignat*}
Note that type substitution $[\sigma]\;A$ is needed, because $A$ is well-formed under the context $\Delta$ instead of $\Gamma$.
As substitution calculus is the initial cwf with a constant type $\UU$, substitutions also have the identity substitution $\idS$ and composition $\blank;\blank$
\begin{alignat*}{3}
  \idS & : \implicit{\Gamma}\;\Sub\;\Gamma\;\Gamma \\
  \blank;\blank & : \implicit{\Gamma, \Delta, \Theta}\;\Sub\;\Gamma\;\Delta \to \Sub\;\Delta\;\Theta \to \Sub\;\Gamma\;\Theta \\
\end{alignat*}
satisfying certain laws (introduced later).
Context comprehension is given by projections from \emph{non-empty} substitutions $\Sub\;\Gamma\;(\Delta, A)$:
\begin{alignat*}{3}
  \pi_1 & : \implicit{\Gamma, \Delta, A}\;\Sub\;\Gamma\;(\Delta, A) \to \Sub\;\Gamma\;\Delta \\
  \pi_2 & : \implicit{\Gamma, \Delta, A}\;(\sigma : \Sub\;\Gamma\;(\Delta, A)) \to \Tm\;\Gamma\;([ \pi_1\,\sigma ]\; A)
\end{alignat*}
where $\pi_1\,\sigma$ and $\pi_2\,\sigma$ can be intuitively understood as the \emph{tail} and the \emph{head} of a non-empty substitution $\sigma : \Sub\;\Gamma\;(\Delta, A)$ respectively.
Again, type substitution $[\pi_1\,\sigma]\,A$ is needed for~$\pi_2$, because $A$ is well-formed under the context $\Delta$ instead of $\Gamma$.
Finally, we also have term substitution $[\sigma]\,t$ whose type is $[\sigma]\,A$ for a term $t : \Tm\;\Delta\;A$
\begin{alignat*}{3}
  [\blank] \blank & : \implicit{\Gamma,\Delta, A}\;(\sigma : \Sub\;\Gamma\;\Delta) \to \Tm\;\Delta\;A \to \Tm\;\Gamma\;([\sigma]\; A)
\end{alignat*}
where the symbol $[\blank]\blank$ is overloaded.

@@

Laws for substitutions are stipulated using equality constructors.
\begin{alignat*}{5}
  \mathsf{idr}    & : \implicit{\Gamma, \Delta, \sigma} && {\sigma ; \idS_{\Delta}} && =^{\Sub\,\Gamma\,\Delta} && {\sigma} \\
  \mathsf{idl}    & : \implicit{\Gamma, \Delta, \sigma} && {\idS_{\Gamma} ; \sigma} && =^{\Sub\,\Gamma\,\Delta} && {\sigma} \\
  ;\text{-}\mathsf{assoc} & : \implicit{\Gamma, \Delta, \Theta, \Xi, \sigma, \tau, \gamma} && (\sigma ; \tau) ; \gamma && =^{\Sub\,\Gamma\,\Theta} &&  \sigma ; (\tau ; \gamma) \\
  \mathsf{concat} & : \implicit{\Gamma, \Delta, \Theta, \sigma, \tau, A, t} && \sigma ; (\tau , t)      && =^{\Sub\,\Gamma\,(\Theta, A)} &&  (\sigma ; \tau) , \alert{\transfib{\Tm\,\Gamma}{[;]_{\Ty}^{-1}}{\color{black}[ \sigma ] t}} \\
  \emptyctx\eta   & : \implicit{\Gamma, A, \sigma} && \sigma                   && =^{\Sub\,\Gamma\,\emptyctx} & \emptysub \\
  \pi\eta         & : \implicit{\Gamma, \Delta, \sigma} && \sigma                   && =^{\Sub\,\Gamma\,(\Delta, A)} &&  (\pi_1 \sigma, \pi_2 \sigma) \\
\end{alignat*}
In particular, $\blank;\blank$ acts like a list concatenation, so conceptually $\sigma; (\tau, t)$ is equal to $(\sigma; \tau), t$ whereas the term $t$ is of type $\Tm\;\Delta\;([\tau]\;A)$. 
Moving the term $t$ from the left-hand side to the right-hand side requires us to apply substitution $\sigma$ to the term $t$ to obtain an inhabitant of $\Tm\;\Gamma\;([\sigma][\tau]\,A)$.
By definition of the substitution extension, however, $[\sigma]\;t$ requires to be $\Tm\;\Gamma\;([\sigma; \tau]\,A)$ instead.
Therefore, we need to transport $[\sigma]\;t$ explicitly along $[;]_{\Ty}^{-1}\colon [\sigma]\,[\tau]\;A = [\sigma;\tau]\,A$ to obtain a term of type $\Tm\;\Gamma\;([\sigma;\tau]\;A)$.

@@

Similarly, for term substitution, terms are transported along the corresponding rules:
\begin{alignat*}{5}
  [\idS]_t         & : \implicit{\Gamma, A, t} && {[\idS]\,t}         && =^{\Tm\;\Gamma}_{\alert{[\idS]_\Ty}}  && t \\
  [;]_t            & : \implicit{\Gamma, \Delta, \Theta, \sigma, \tau, t} && {[\sigma ; \tau]\,t} && =^{\Tm\;\Gamma}_{\alert{[;]_{\Ty}}}   && {[ \sigma ]\,[ \tau ]\,t}
\end{alignat*}
equivalent to the following (homogeneous) identities
\[
  \transfib{\Tm\;\Gamma}{[\idS]_{\Ty}}{[\idS]\,t} =^{\Tm\,\Gamma\,A} t
  \quad\text{and}\quad
  \transfib{\Tm\;\Gamma}{[;]_{\Ty}}{[\sigma;\tau]\,t} =^{\Tm\,\Gamma\,([\sigma]\,[\tau]A)} [\sigma]\,[\tau]\,t
\]
respectively.
Constructors introduced so far complete the definition of substitution calculus.

@@

To retain the intuitive way of reasoning such as \eqref{eq:pi2-comp-proof} \emph{formally}, we would like to make type substitution rules definitional, since the root cause is that type substitution as a constructor cannot compute so that type mismatches occur after substitution.

One possibility is to define substitution calculus as a quotient inductive-inductive-\emph{recursive} type whereas type substitution as a simultaneous recursive function computes substitution at the meta-level.
Now that $[ \idS ]_{\Ty}\;A = A$ and $[ \sigma ; \tau ]_{\Ty}\;A = [ \sigma ]_{\Ty}\;([ \tau ]_{\Ty}\;A)$ along do not make $[\blank]_{\Ty}\blank$ a function, we may define it by
\begin{alignat*}{3}
[\blank]_{\Ty} \blank &: \implicit{\Gamma, \Delta} \; \Sub\;\Gamma\;\Delta \to \Ty\;\Delta \to \Ty\;\Gamma \\
[ \sigma ]_{\Ty}\;\UU & = \UU
\end{alignat*}
However, this definition is \emph{not} inductive-recursive in the sense of Dybjer and Setzer~\cite{Dybjer2000,Dybjer2003}: in their schema, the codomain of a simultaneous recursive function does not refer to any inductive type being defined.
While its general schema remains underdeveloped, this definition is accepted in proof assistants such as \Agda and used in other formalisation of type theory~\cite{Danielsson2006}.
Still, although $[\idS]_{\Ty}\,A$ and $[\sigma; \tau]_{\Ty}\,A$ can be proved equal to $A$ and $[\sigma]_{\Ty}\,[\tau]_{\Ty}\,A$ respectively, they are not definitional but propositional.
This definition fails to eliminate any transport in the definition of substitution calculus.

Our second attempt is to define type substitution with additional rules:
\begin{alignat}{3}
[\blank]_{\Ty} \blank &: \implicit{\Gamma, \Delta} \; \Sub\;\Gamma\;\Delta \to \Ty\;\Delta \to \Ty\;\Gamma \nonumber \\
[ \sigma ]_{\Ty}       \;\UU  & = \UU                                   \label{eq:type-sub-at2-1} \\
[ \idS ]_{\Ty}         \;A    & = A                                     \label{eq:type-sub-at2-2} \\
[ \sigma ; \tau ]_{\Ty}\;A    & = [ \sigma ]_{\Ty}\;([ \tau ]_{\Ty}\;A) \label{eq:type-sub-at2-3}
\end{alignat}
However, this definition does \emph{not} reduce \eqref{eq:type-sub-at2-2} or \eqref{eq:type-sub-at2-3} definitionally in a type-theoretic proof assistant.
By design, definitions by pattern matching will be translated to a form that corresponds to the \emph{eliminator} either using \emph{first-match semantics}~\cite{Cockx2020a} or an explicit construct for pattern matching, so only the eliminator for $\Ty\;\Delta$ will be considered in the above definition.
\LT[noinline]{How about \Coq and \Lean?}
Using the first-match semantics, the above definition amounts to the following definition:
\begin{alignat*}{3}
[\blank]_{\Ty} \blank &: \implicit{\Gamma, \Delta} \; \Sub\;\Gamma\;\Delta \to \Ty\;\Delta \to \Ty\;\Gamma \\
[ \sigma ]_{\Ty}\;\UU  & \reduce \UU
\end{alignat*}
where $\reduce$ is used to emphasise the reduction (which entails the definitional equality): we have done nothing different from our first attempt! 
%As a result, neither  are definitional.
We may swap \eqref{eq:type-sub-at2-2} with \eqref{eq:type-sub-at2-2} and \eqref{eq:type-sub-at2-3} to make \eqref{eq:type-sub-at2-2} and \eqref{eq:type-sub-at2-3} definitional, but then \eqref{eq:type-sub-at2-1} will be translated to 
\begin{alignat*}{3}
[ \emptysub ]_{\Ty}\;\UU         & \reduce \UU \\
[ \sigma, t ]_{\Ty}\;\UU         & \reduce \UU \\
[ \pi_1\,\sigma ]_{\Ty}\;\UU     & \reduce \UU,
\end{alignat*}
making $[\sigma]\;\UU = \UU$ propositional instead.
Not being able to reduce $[ \sigma ]_{\Ty}\;\UU$ to $\UU$ would make our later extension with a type $\El\,a$ for $a : \Tm\;\Gamma\;\UU$ require another transport in the definition.
Likewise, for any extension with a type former such as $\Pi$-types, structural rules would still be propositional.
That is, using the standard semantics of function definition, we are not able to make type substitution definitional on \emph{both} $\Sub\;\Gamma\;\Delta$ and $\Ty\;\Delta$.

To ensure that type substitution is definitional for all \emph{rules}, we turn to a definition by overlapping patterns (\Cref{sec:meta:overlapping}) instead to make each clause definitional. 
In addition, we can include other derived rules such as $[\pi_1(\sigma, t)]_{\Ty}\;A = [ \sigma]_{\Ty}\;A$ as long as it is locally confluent and terminating.
We then end up with the following definition:
\begin{alignat}{3}
[\blank]_{\Ty} \blank            &: \implicit{\Gamma, \Delta} \; \Sub\,\Gamma\,\Delta \to \Ty\,\Delta \to \Ty\,\Gamma \nonumber \\
[ \idS ]_{\Ty}\;A                & \reduce A \label{eq:def-type-subst-1}\\
[ \sigma ; \tau ]_{\Ty}\;A       & \reduce [ \sigma ]_{\Ty}\;([ \tau ]_{\Ty}\;A) \label{eq:def-type-subst-2}\\
[ \pi_1(\sigma, t) ]_{\Ty}\;A    & \reduce [\sigma]_\Ty\;A                       \label{eq:def-type-subst-3} \\
[ \pi_1(\sigma; \tau) ]_{\Ty}\;A & \reduce [\sigma]_\Ty\;([\pi_1\tau]_\Ty\;A)    \label{eq:def-type-subst-4} \\
[ \sigma ]_{\Ty}\;\UU            & \reduce \UU                                   \label{eq:def-type-subst-5}
\end{alignat}
We first check that this definition of type substitution is valid.\footnote{%
\Agda implements a local confluence checker with the experimental option \text{-{}-local-confluence-check} which is turned on globally in our formalisation, but we do not trust experimental features blindly.}
\begin{proposition}[Local confluence] \label{prop:local-confluence-1}
  Type substitution $[\blank]_{\Ty}\blank$ is locally confluent.
  \danger
\end{proposition}
\begin{proof}
  To show the local confluence, we consider all local peaks, i.e.\ each combination of \eqref{eq:def-type-subst-5} with \eqref{eq:def-type-subst-1}--\eqref{eq:def-type-subst-4},
  \[
    [\idS]_{\Ty}\;\UU, \qquad [\sigma;\tau]_{\Ty}\;\UU, \qquad [\pi_1(\sigma, t)]_{\Ty}\;\UU, \qquad\text{and}\qquad [\pi_1(\sigma; \tau)]_{\Ty}\;\UU
  \]
  and show that each applicable clause does result in the same term $\UU$.

  For example, $[\sigma;\tau]_{\Ty}\;\UU \reduce [\sigma]_{\Ty}\;([\tau]_{\Ty}\;\UU) \reduce [\sigma]_{\Ty}\;\UU \reduce \UU$ using \eqref{eq:def-type-subst-2} as the first reduction and $[\sigma;\tau]_{\Ty}\;\UU \reduce \UU$ using \eqref{eq:def-type-subst-5} as the first reduction.
  It is easy to see that remaining local peaks reduce to the same term $\UU$.
